---
title: "Codebook"
author: "phbehre"
date: "March 26, 2016"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The following file describes the activities perfromed as part of the Getting and Cleaning Data Course Project and explains the run_analysis.R script. The script use data collected from the accelerometers from the Samsung Galaxy S smartphone. The data was collected in the context of the UCI Human Activity Recognition Using Smartphones Data Set. Please see more details on the project and the related dataset [here](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones).

Please find the dataset used via the following link [Project Dataset](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip). After extracting the zip file, you can find detailed information  about the structure of the dataset in /UCI HAR Dataset/README.txt.

run_analysis.R performs the following activities:

## 0. Libraries used
The script run_analysis.R uses the following libraries

```{r}
library(data.table)
```


## 1. Download the data 
Downloads the original data set from the project [Project Dataset](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip). 

```{r}
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(fileUrl,destfile="./data/UCI_HAR_dataset.zip",method="curl")
```

## 2. Unzip the data file
Unzips the file "UCI HAR Dataset.zip"

```{r}
unzip(zipfile="./data/UCI_HAR_dataset.zip",exdir="./data")
```

the folder should contain the follwoing files:

```{r}
path_data_files <- file.path("./data" , "UCI HAR Dataset")
files<-list.files(path_data_files, recursive=TRUE)
files
```

## 3. read data from files into data frames

```{r}
featureNames <- read.table("./data/UCI HAR Dataset/features.txt")
activityLabels <- read.table("./data/UCI HAR Dataset/activity_labels.txt", header = FALSE)

subjectTrain <- read.table("./data/UCI HAR Dataset/train/subject_train.txt", header = FALSE)
activityTrain <- read.table("./data/UCI HAR Dataset/train/y_train.txt", header = FALSE)
featuresTrain <- read.table("./data/UCI HAR Dataset/train/X_train.txt", header = FALSE)

subjectTest <- read.table("./data/UCI HAR Dataset/test/subject_test.txt", header = FALSE)
activityTest <- read.table("./data/UCI HAR Dataset/test/y_test.txt", header = FALSE)
featuresTest <- read.table("./data/UCI HAR Dataset/test/X_test.txt", header = FALSE)
```

## 4. review the properties of the created variables

```{r}
str(subjectTrain)
str(activityTrain) 
str(featuresTrain)
str(subjectTest) 
str(activityTest)
str(featuresTest)
```

# Merge the training and the test sets to create one data set

## 5. Merge th data in training and test data sets corresponding to subject, activity and features.

```{r}
subject <- rbind(subjectTrain, subjectTest)
activity <- rbind(activityTrain, activityTest)
features <- rbind(featuresTrain, featuresTest)
```

## 6. set meaningful column names

```{r}
colnames(features) <- t(featureNames[2])

colnames(activity) <- "Activity"
colnames(subject) <- "Subject"
```

## 7. Merge complete data set
merge the complete data set and review the head data of the variable "completeData"
```{r}
completeData <- cbind(features,activity,subject)
head(completeData)
```

# Extracts only the measurements on the mean and standard deviation for each measurement.

## 8. Compelte the dataset and extract only measurements on mean and standard deviation for the measurement.
Identify and extract the column indices that have the mean and std in the name

```{r}
columnsWithMeanSTD <- grep(".*Mean.*|.*Std.*", names(completeData), ignore.case=TRUE)
```

Activity and Subject columns need to be added to the list.View the dimensions of "completeData".

```{r}
requiredColumns <- c(columnsWithMeanSTD, 562, 563)
dim(completeData)
```

Generate the completed dataset "Data". View dimensions of the "Data".

```{r}
Data <- completeData[,requiredColumns]
dim(Data)
```
# Uses descriptive activity names to name the activities in the data set

## 9. Uses descriptive activity names to name the activities in the data set

```{r}
Data$Activity <- as.character(Data$Activity)
for (i in 1:6){
  Data$Activity[Data$Activity == i] <- as.character(activityLabels[i,2])
}
```

Factor the activity variable after updating the variable names

```{r}
Data$Activity <- as.factor(Data$Activity)
```

# Appropriately labels the data set with descriptive variable names. 

## 10. Replace the varibale names with more descriptive names

```{r}
# Appropriately labels the data set with descriptive variable names
names(Data)<-gsub("Acc", "Accelerometer", names(Data))
names(Data)<-gsub("Gyro", "Gyroscope", names(Data))
names(Data)<-gsub("BodyBody", "Body", names(Data))
names(Data)<-gsub("Mag", "Magnitude", names(Data))
names(Data)<-gsub("^t", "Time", names(Data))
names(Data)<-gsub("^f", "Frequency", names(Data))
names(Data)<-gsub("tBody", "TimeBody", names(Data))
names(Data)<-gsub("-mean()", "Mean", names(Data), ignore.case = TRUE)
names(Data)<-gsub("-std()", "STD", names(Data), ignore.case = TRUE)
names(Data)<-gsub("-freq()", "Frequency", names(Data), ignore.case = TRUE)
names(Data)<-gsub("angle", "Angle", names(Data))
names(Data)<-gsub("gravity", "Gravity", names(Data))
```

Review the names after updating

```{r}
names(Data)
```

# From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

## 11. Set "subject" as a factor variable

```{r}
Data$Subject <- as.factor(Data$Subject)
Data <- data.table(Data)
```

## 12. Create tidy dataset and write it to file

```{r}
tidyDataset <- aggregate(. ~Subject + Activity, Data, mean)
tidyDataset <- tidyDataset[order(tidyDataset$Subject,tidyDataset$Activity),]
write.table(tidyDataset, file = "UCI_HAR_tidy_dataset.txt", row.names = FALSE)
```